Linking watershed nutrient loading to estuary water quality with Generalized Additive Models

### 2023-08-08 Schramm-2023-06-PeerJ/Schramm-2023-08-PeerJ.tex
incorporate new estuary models

\title{Linking watershed nutrient loading to estuary water quality with
semiparametric models}

@@

\caption{\label{tab:table2}Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.}
\centering
\begin{threeparttable}
\begin{tabular}[t]{lllll>{\raggedright\arraybackslash}p{3cm}rr}
\toprule
Station ID & Parameter & Mean & SD & N & Method & AWRL\textsuperscript{a} & Criteria\textsuperscript{b}\\
\midrule
 & TP (mg/L) & 0.11 & 0.05 & 47 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13383} & DO (mg/L) & 7.22 & 1.35 & 55 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.08 & 0.03 & 51 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13384} & DO (mg/L) & 7.51 & 1.32 & 54 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.13 & 0.06 & 50 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13563} & DO (mg/L) & 7.91 & 1.34 & 56 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item[a] Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.
\item[b] The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes. NO\textsubscript{x} and TKN parameters are not currently screened by the state and do not have an associated screening level.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

where \(\mu\) is the conditional expected response (nutrient
concentration), \(g()\) is the log link, and response variable was
modeled as Gamma distributed with mean \(\mu\) and scale \(\lambda\).
\(f_1(ddate)\) is decimal date smoothed with a thin-plate regression
spline, \(f_2(yday)\) is the numeric day of year smoothed with a cyclic
cubic regression spline, \(f_3(Q)\) is mean daily inflow (the combined
measurements from Lavaca River and Navidad River) and \(f_4(Load)\) is
the total NO\textsubscript{3} or TP watershed load. The set of models
specified for each water quality response are in Table~\ref{tab:table3}.

@@

Because streamflow and nutrient loads are tightly correlated, freshwater
inflow can mask signals due to changes in nutrient loads alone.
Following the methodology implemented by
\textcite{murphyNutrientImprovementsChesapeake2022}, both freshwater
inflow and nutrient loads were prepossessed to account for season and
streamflow respectively. Raw freshwater inflow values were replaced by
seasonally adjusted log transformed inflow obtained from the residuals
of a GAM model fit between season(day of year) and log transformed daily
freshwater inflow. Raw nutrient loads were replaced with flow-adjusted
values obtained from the residuals of a GAM model relating log
transformed NO\textsubscript{3} or TP loads to log transformed daily
inflow. Response residuals from the respective GAM models were used as
\emph{Q} and \emph{Load} in Eq.~\ref{eq6} and Eq.~\ref{eq7}.

@@

\caption{\label{tab:table4}Estuary GAM AIC\textsubscript{c} values and associated model probabilities (in parenthesis). Models with the highest probability for each site and water quality parameter combination are bolded and italicized for emphasis.}
\centering
\begin{tabular}[t]{ll>{}l>{}l>{}l}
\toprule
Parameter & Site & Temporal & Inflow & Inflow + Load\\
\midrule
 & TCEQ-13383 & -152.1 (0.03) & -156.1 (0.24) & \em{\textbf{-158.2 (0.72)}}\\

 & TCEQ-13384 & -194.4 (0.03) & \em{\textbf{-200.2 (0.49)}} & -200.2 (0.49)\\

\multirow{-3}{*}{\raggedright\arraybackslash TP} & TCEQ-13563 & -145.3 (0) & -156.6 (0.41) & \em{\textbf{-157.3 (0.59)}}\\
\cmidrule{1-5}
 & TCEQ-13383 & -218.9 (0) & \em{\textbf{-244.8 (0.5)}} & -244.8 (0.5)\\

 & TCEQ-13384 & -263.4 (0) & -311.7 (0.48) & \em{\textbf{-311.9 (0.52)}}\\

\multirow{-3}{*}{\raggedright\arraybackslash NO\textsubscript{x}} & TCEQ-13563 & -175.1 (0) & \em{\textbf{-190.2 (0.5)}} & -190.2 (0.5)\\
\cmidrule{1-5}
 & TCEQ-13383 & 279.7 (0.18) & \em{\textbf{278.1 (0.41)}} & 278.1 (0.41)\\

 & TCEQ-13384 & \em{\textbf{268.2 (0.33)}} & 268.2 (0.33) & 268.2 (0.33)\\

\multirow{-3}{*}{\raggedright\arraybackslash Chlorophyll-\textit{a}} & TCEQ-13563 & 289.5 (0.08) & \em{\textbf{286.1 (0.46)}} & 286.1 (0.46)\\
\cmidrule{1-5}
 & TCEQ-13383 & \em{\textbf{42.2 (0.66)}} & 43.5 (0.34) & -\\

 & TCEQ-13384 & \em{\textbf{34.3 (0.57)}} & 34.8 (0.43) & -\\

\multirow{-3}{*}{\raggedright\arraybackslash TKN} & TCEQ-13563 & 31.1 (0.22) & \em{\textbf{28.7 (0.78)}} & -\\
\cmidrule{1-5}
 & TCEQ-13383 & \em{\textbf{146.4 (0.34)}} & 146.4 (0.34) & 146.5 (0.32)\\

 & TCEQ-13384 & \em{\textbf{135.9 (0.47)}} & 137 (0.27) & 137 (0.27)\\

\multirow{-3}{*}{\raggedright\arraybackslash DO} & TCEQ-13563 & 138.3 (0.25) & \em{\textbf{137.2 (0.43)}} & 137.8 (0.32)\\
\bottomrule
\end{tabular}
\end{table}

>>>>

\title{Linking watershed nutrient loading to estuary water quality with
Generalized Additive Models}

@@

\caption{\label{tab:table2}Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}llll>{\raggedright\arraybackslash}p{2.5cm}rr}
\toprule
Station ID & Parameter & Mean & SD & N & Method & AWRL\textsuperscript{a} & Criteria\textsuperscript{b}\\
\midrule
 & TP (mg/L) & 0.11 & 0.05 & 47 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{2cm}{\raggedright\arraybackslash TCEQ-13383, Upper-Bay} & DO (mg/L) & 7.22 & 1.35 & 55 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.08 & 0.03 & 51 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{2cm}{\raggedright\arraybackslash TCEQ-13384, Lower-Bay} & DO (mg/L) & 7.51 & 1.32 & 54 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.13 & 0.06 & 50 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{2cm}{\raggedright\arraybackslash TCEQ-13563, Mid-Bay} & DO (mg/L) & 7.91 & 1.34 & 56 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item[a] Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.
\item[b] The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes. NO\textsubscript{x} and TKN parameters are not currently screened by the state and do not have an associated screening level.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

where \(\mu\) is the conditional expected response (nutrient
concentration), \(g()\) is the log link, and response variable was
modeled as Gamma distributed with mean \(\mu\) and scale \(\lambda\).
\(f_1(ddate)\) is decimal date smoothed with a thin-plate regression
spline, \(f_2(yday)\) is the numeric day of year smoothed with a cyclic
cubic regression spline. Prior work has shown that many water quality
parameters may have lagged effects lasting days or even months following
storms and large discharge events
\autocite{mooneyWatershedExportEvents2012a,wetzExtremeFutureEstuaries2013,bukaveckasInfluenceStormEvents2020,walkerTimescalesMagnitudeWater2021}.
\textcite{walkerTimescalesMagnitudeWater2021} showed the impact of large
flow events like Hurricane Harvey had somewhat short water quality
impacts in Lavaca Bay and adjacent estuary systems, ranging somewhere
around 20 days to a few months. We incorporate this lag effect by using
cumulative totals for \(f_3(Q)\) and \(f_4(Load)\). \(f_3(Q)\) is the
cumulative of the previous 20 days of inflow (combined measurements from
Lavaca River and Navidad River) and \(f_4(Load)\) is the cumulative of
the previous 20 days total NO\textsubscript{3} or TP watershed load. The
set of models specified for each water quality response are in
Table~\ref{tab:table3}.

@@

Because streamflow and nutrient loads are tightly correlated, freshwater
inflow can mask signals due to changes in nutrient loads alone.
Following the methodology implemented by
\textcite{murphyNutrientImprovementsChesapeake2022}, both freshwater
inflow and nutrient loads were prepossessed to account for season and
streamflow respectively. Cumulative inflow values were replaced by
seasonally adjusted cumulative inflow obtained from the residuals of a
GAM model fit between season(day of year) and log transformed cumulative
inflow. Nutrient loads utilized the flow-normalized loads estimated in
the previous section.

@@

\caption{\label{tab:table4}Estuary GAM AIC\textsubscript{c} values and associated model probabilities (in parenthesis). Models with the highest probability for each site and water quality parameter combination are bolded and italicized for emphasis.}
\centering
\begin{tabular}[t]{ll>{}l>{}l>{}l}
\toprule
Parameter & Site & Temporal & Inflow & Inflow + Load\\
\midrule
 & Upper-Bay & -152.1 (0.02) & -155.1 (0.07) & \em{\textbf{-160.2 (0.91)}}\\

 & Mid-Bay & -145.3 (0.01) & -150 (0.12) & \em{\textbf{-154 (0.87)}}\\

\multirow{-3}{*}{\raggedright\arraybackslash TP} & Lower-Bay & -194.4 (0.39) & \em{\textbf{-194 (0.31)}} & -194 (0.31)\\
\cmidrule{1-5}
 & Upper-Bay & -218.9 (0) & \em{\textbf{-241.7 (0.5)}} & -241.7 (0.5)\\

 & Mid-Bay & -175.1 (0.01) & \em{\textbf{-183.8 (0.5)}} & -183.8 (0.5)\\

\multirow{-3}{*}{\raggedright\arraybackslash NO\textsubscript{x}} & Lower-Bay & -263.4 (0) & -298.1 (0.5) & \em{\textbf{-298.1 (0.5)}}\\
\cmidrule{1-5}
 & Upper-Bay & 279.7 (0.24) & \em{\textbf{279.6 (0.26)}} & 278.2 (0.5)\\

 & Mid-Bay & 289.5 (0.37) & \em{\textbf{289.5 (0.38)}} & 290.3 (0.25)\\

\multirow{-3}{*}{\raggedright\arraybackslash Chlorophyll-\textit{a}} & Lower-Bay & \em{\textbf{268.2 (0.03)}} & 262.7 (0.48) & 262.7 (0.48)\\
\cmidrule{1-5}
 & Upper-Bay & \em{\textbf{42.2 (0.5)}} & 42.2 (0.5) & -\\

 & Mid-Bay & 31.1 (0.5) & \em{\textbf{31.1 (0.5)}} & -\\

\multirow{-3}{*}{\raggedright\arraybackslash TKN} & Lower-Bay & \em{\textbf{34.3 (0.5)}} & 34.3 (0.5) & -\\
\cmidrule{1-5}
 & Upper-Bay & \em{\textbf{146.4 (0.37)}} & 146.8 (0.29) & 146.5 (0.34)\\

 & Mid-Bay & 138.3 (0.17) & \em{\textbf{136.4 (0.42)}} & 136.4 (0.42)\\

\multirow{-3}{*}{\raggedright\arraybackslash DO} & Lower-Bay & \em{\textbf{135.9 (0.04)}} & 130.6 (0.48) & 130.6 (0.48)\\
\bottomrule
\end{tabular}
\end{table}
